### 2020-06-21 chapters/background.tex
EMA histograms

For the scope of this Thesis, we will work under a tuple-based true sliding window framework. Despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.

@@

where $\tau$ is the half-life of the EMA. The half-life $\tau$ controls how fast older events are forgotten as the smoothing factor depends on it. The half-life is the amount of time that it takes for an element to see its contribution or weight reduced to half. For instance, with $\tau = 1s$ we have $\lambda = 0.5$, which means for each 1s step we reduce the weights applied to each event by half.

>>>>

For the scope of this Thesis, we will work under a tuple-based true sliding window framework. However, as explained in Section \ref{sec:ft-monitoring}, despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.

@@

where $\tau$ is the half-life of the EMA. The half-life $\tau$ controls how fast older events are forgotten as the smoothing factor depends on it. The half-life is the amount of time that it takes for an element to see its contribution or weight reduced to half. For instance, with $\tau = 1s$ we have $\lambda = 0.5$, which means for each 1s step we reduce the weights applied to each event by half. Note that it is also possible to express a tuple-based half-life, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half (\textit{e.g.} $\tau = 1$ tuple, $\lambda = 0.5$, for each new tuple reduce all weights by half).

### 2020-06-21 chapters/my-work.tex
EMA histograms

Using point outlier detection methods, the main idea would be to keep track of the rate of point outliers detected. For instance, keep track of how many outliers were alerted in the past hour. The decision to alert for a data pattern shift of the system would be made when comparing this rate of point outliers detected in the past hour with a user-defined threshold $\alpha$. On the other hand, using subsequence outlier detection methods would allow us to report the whole sliding window as an outlier.

@@

In our lightweight real-time solution we need constant time complexity to process events one by one in real-time. Additionally, because our system is not mission-critical, we want a solution as lightweight as possible, which means something that grows linearly with window size is not a good fit. Therefore, most of the outlier detection methods presented become unfit for our use due to high memory consumption and time complexity. Some of the methods shown also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period. In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream. While in some scenarios this is a much-needed feature, in our use-case it is not. How clear the alerts and insights retrieved from the methods are, is also an important property. Some outlier detection methods used Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.

@@

\section{Proposed method: Feature Distribution Monitoring}

@@

To compare each feature's reference and sliding window histogram aggregation we need a measure of divergence. We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein. \textcolor{red}{defer the why to later section} Given a divergence value for the two aggregations (reference and sliding window ones) we need a measure of normal statistical fluctuations. Since we assume a reference period as the normal state we can find out the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis. This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand picked threshold for the divergence value. \textcolor{red}{defer details to later section}

We devise a two-phase method. The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section ??). The second phase focus on efficiently updating the sliding window histogram for each feature and periodically performs a divergence test between the static reference aggregation and the sliding one, alerting for changes if need be. The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.

@@

\subsection{Exponential Moving Average based histogram} \label{sec:ema-hist}
The sliding window histogram we use is an Exponential Moving Average (EMA) based histogram. An EMA-like histogram is essentially a collection of EMA-counts as described in Section \ref{sec:emas}.

.....

\subsection{Batch analysis phase}
\textcolor{red}{TODO: add some pseudocode}

@@

\subsection{Streaming phase}
\textcolor{red}{TODO: add some pseudocode}

@@

>>>>

Using point outlier detection methods, the main idea is to keep track of the rate of point outliers detected. For instance, keep track of how many outliers were alerted in the past hour. The decision to alert for a data pattern shift of the system would be made when comparing this rate of point outliers detected in the past hour with a user-defined threshold $\alpha$. On the other hand, using subsequence outlier detection methods would allow us to report the whole sliding window as an outlier.

@@

In our lightweight real-time solution we need constant time complexity to process events one by one in real-time. Because our system is not mission-critical, we also want the solution to be as lightweight as possible, which means techniques that grow linearly with window size are not a good fit. Therefore, most of the outlier detection methods presented become unfit for our use due to high memory consumption and time complexity. Some of the methods discussed also suffer from what we call \textit{"memory loss"}: they lose track of the reference window or normality period. In other words, the algorithms are online learners, meaning they adapt to the underlying statistics of the data stream. While in some scenarios this is a much-needed feature, in our use-case it is not. How clear the alerts and insights retrieved from the methods are, is also an important property. Some outlier detection methods use Machine Learning (ML) techniques to produce alerts, which makes it harder to explain to a system administrator why an alert was raised and what changed.

@@

\section{Proposed method: Feature Distribution Monitoring} \label{sec:ft-monitoring}

@@

To compare each feature's reference and sliding window histogram aggregation we need a measure of divergence. We chose to use the Jensen–Shannon Divergence (JSD) \cite{JSD} but other statistical tests are possible such as Kolmogorov-Smirnov or Wasserstein. A more thorough analysis on some statistical tests and why we chose to use the JSD is performed in Section \ref{sec:whyjsds}. Given a divergence value for the two aggregations (reference and sliding window ones) we need a measure of normal statistical fluctuations. Since we assume a reference period as the normal state we can find out the normal distribution for the divergence by sampling smaller windows of data from the reference period in a batch analysis. This allows us to define a threshold based on percentiles, which gives an estimate for the probability of a certain value occurring, rather than using a constant hand picked threshold for the divergence value. \textcolor{red}{defer details to later section}

We devise a two-phase method. The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section \ref{sec:batch-phase}). The second phase (detailed in Section \ref{sec:stream-phase}) focus on efficiently updating the sliding window histogram for each feature and periodically performs a divergence test between the static reference aggregation and the sliding one, alerting for changes if need be. The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.

@@

\subsection{Exponential Moving Average based histogram} \label{sec:ema-hist}
As stated, we compute histogram aggregations for the reference and sliding windows. The computation of the reference histogram is done once in batch and hence an exact histogram can be computed. However, the sliding window histogram is an EMA-like histogram to ensure that we are able to encode the feature distribution with constant space complexity and maintain it in real-time.

In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they emulate a sliding window without actually storing window elements....

\textcolor{red}{TODO: complete this after EMAs in background are corrected}

\subsection{Batch analysis phase} \label{sec:batch-phase}
\textcolor{red}{TODO: add some pseudocode}

@@

\subsection{Streaming phase} \label{sec:stream-phase}
\textcolor{red}{TODO: add some pseudocode}

@@

\subsubsection*{Distribution Divergence Statistical Tests} \label{sec:whyjsds}

We used the Jensen-Shannon Divergence (JSD) but....

\textcolor{red}{Talk about wassertein, kolomogrov, why JSD...}

### 2020-06-21 chapters/problem-statement.tex
EMAs draft
EMA histograms

@@

\section{State of the art issues}
Aggregations, specifically sliding window ones, were introduced in Chapter \ref{chap:background} along with several windowing techniques. In Chapter \ref{chap:sota} we presented state of the art sliding window aggregation algorithms as well as outlier detection methods. However, the reviewed methods were deemed unfit for our use-case because none of them simultaneously showed all of the below characteristics:

>>>>

\textcolor{red}{INCOMPLETE}

@@

\section{State of the art issues}
Aggregations, specifically sliding window ones, were introduced in Chapter \ref{chap:background} along with several windowing techniques required to monitor streaming data patterns in real-time. In Chapter \ref{chap:sota} we presented state of the art sliding window aggregation algorithms as well as outlier detection methods. However, the reviewed methods were deemed unfit for our use-case because none of them simultaneously showed all of the below characteristics:

### 2020-06-21 chapters/sota.tex
EMA histograms

Yang et. al propose a method for incremental detection of density-based point outliers \cite{Yang-Neighbor-Based-Pattern-Detection}, specifically for sliding windows scenarios. The authors propose Extra-N, an algorithm that detects point outliers based on the number of neighbors it has. The authors make use of the \textit{“predictability"} property of the expiration of existing objects. In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point on arrival --- \textit{i.e.} the future windows it will belong to. This technique allows the efficient maintenance of neighborship counts by pre-handling the eventual expiring effect of new data points.

@@

\subsection*{Applicability to our Hypothesis}
Bloom filters exhibit constant space and time complexity making it great methods to test our hypothesis. Their false positive rate is tolerable in most scenarios and can be controlled by tweaking the values of \textit{m} and \textit{k}.

>>>>

Yang et. al propose a method for incremental detection of density-based point outliers \cite{Yang-Neighbor-Based-Pattern-Detection}, specifically for sliding windows scenarios. The authors propose Extra-N, an algorithm that detects point outliers based on the number of neighbors it has. The authors make use of the \textit{“predictability"} property of the expiration of existing objects. In other words, given a window with a fixed slide size, they predetermine the \textit{life-span} of any data point on arrival --- \textit{i.e.} the future windows it will belong to.

@@

\subsection*{Applicability to our Hypothesis}
Bloom filters exhibit constant space and time complexity making them possible methods to test our hypothesis. Their false positive rate is tolerable in most scenarios and can be controlled by tweaking the values of \textit{m} and \textit{k}. However, we need to evict old events, and for that reason, bloom filters are of no use to us (unless we find sliding window implementations).
